# Different ways to summarise

```{r, message=FALSE}
library(tidyverse)
```

Let's use `mtcars` data, but first we need to convert two numeric variables into factor:

```{r}
# fix some numeric variables into factor
dt <- mutate( mtcars,
  vs = factor(vs, levels=c(0, 1), labels=c("V-shaped", "straight")),
  am = factor(am, levels=c(0, 1), labels=c("automatic", "manual"))
)

glimpse(dt)
```

## `desc_table`

```{r, message=FALSE}
library(desctable)
```

The library `desctable` provides `desc_table()` function to calculate main descriptive statistics. The output is a new dataframe. You can also change the output using `desc_output()` function.

**Numeric variables**

```{r}
desc_table(dt %>% select(-c(am, vs))) %>% desc_output('pander')
```

**Categorical variables**

```{r}
desc_table(dt %>% select(c(am, vs))) %>% desc_output("pander")
```

See more: [https://cran.r-project.org/web/packages/desctable/vignettes/desctable.html](https://cran.r-project.org/web/packages/desctable/vignettes/desctable.html){target="_blank"}

## `skim`

```{r, message=FALSE}
library(skimr)
```

`skim()` from `skimr` package provides a complete summary separated between numeric and categorical variables. Interestingly, histogram bars are shown in the last column for numeric variables.

```{r}
skim(dt)
```

See more: [https://cran.r-project.org/web/packages/skimr/vignettes/skimr.html](https://cran.r-project.org/web/packages/skimr/vignettes/skimr.html){target="_blank"}

## `tableone`

```{r}
library(tableone)
```

Table 1 is a common name used in biomedical research paper that describes the patient demographics. A package called `tableone` aims to ease the production of this table, and we can use this package to summarise our data.

```{r}
CreateTableOne(data=dt, strata="am")
```

## `tbl_summary`

```{r}
library(gtsummary)
```

The `gtsummary` package provides a rich collection of functions for summarising tables and results from different statistical analyses, e.g. regression, survival analysis, etc.. The simple one, `tbl_summary()`, can generate a beautiful summary table, ready for publication.

```{r, warning=FALSE, message=FALSE}
dt %>%
  tbl_summary(
    by = am,
    label = c(
      mpg ~ "Miles/gallon (US)",
      cyl ~ "Number of cylinders",
      disp ~ "Displacement (cu.in)",
      hp ~ "Gross horsepower",
      drat ~ "Rear axle ratio",
      wt ~ "Weight (1,000 lbs)",
      qsec ~"Quarter mile time",
      vs ~ "Engine type",
      gear ~ "Number of forward gears",
      carb ~ "Number of carburators"
    ),
    statistic = c(all_continuous() ~ "{mean} ± {sd}")
  ) %>%
  add_p() %>%
  separate_p_footnotes() %>%
  add_overall(last = TRUE) %>%
  modify_spanning_header(c("stat_1", "stat_2") ~ "**Transmission**")
```

See more: [https://www.danieldsjoberg.com/gtsummary/index.html](https://www.danieldsjoberg.com/gtsummary/index.html){target="_blank"}
